---
title: "my_vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{my_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(toolbox.e3d)
```

# Intro

The toolbox.e3d was created to perform often occuring tasks done with the soil erosion model EROSION-3D in an easy usable and reproducable way.
It requires a licenced EROSION-3D installation on a Windows system.
Download of this software and licence conditions are available at http://bodenerosion.com/e3d_lizenzen.html.

EROSION-3D is designed as software with Graphical User Interface (GUI) with limited command line support.
The toolbox.e3d uses only the command line interaction, as this is the only way to make the modelling fully reproducable (compare EGU GMD editorial).

# Setup
Once EROSION-3D is installed and activated the interaction with toolbox.e3d can be tested using

```{r}
toolbox.e3d::get_version.E3D()
```

This function checks, if EROSION-3D can be found in the system PATH-Variable.
If it is not present in PATH it searchs for EROSION-3D in the registry and adds it temporarely to PATH for the current R-session.

When succesful it returns the version number of the found EROSION-3D installation.

If this function fails with error message "cannot access e3d instalation" you should check the installtion of EROSION-3D and add the installation folder manually to system PATH.


# determination of calibration parameters

Soil input parameters of EROSION-3D include the two calibration parameters skinfactor and resistance to erosion.
Determination of these two needs to be done to fit model output with runoff/soilloss measurement data obtained in rainfall simulations in field.

Determination is done by automaticly setting up EROSION-3D models of homogeneus rainfall simulation plots and iteration of the relevant calibration parameter until the choosen fitting target is reached.

## fitting targets skinfactor

### cumulative runoff


## fitting targets resistance to erosion


# command line interaction modes
EROSION-3D can be run by command line using the following syntax
"e3d /r C:/E3Dtest/model/run.par"

